Switch equivalence and weak prime domains for fusions

### 2024-01-29 fusione.tex
summary

\tableofcontents
\section{Introduction}

@@

\begin{definition}[Sequential independence]  Let $(\X, \R)$ be a left-linear DPO-rewriting system with $\X$ an $\mathcal{M}$-adhesive category. Let also $\dder{D}\colon G\Mapsto H$ and $\dder{D'}\colon H\Mapsto T$ be the two direct derivations depicted below.

@@

\begin{definition}[Switch]
 Let $(\X, \R)$ be a left-linear DPO-rewriting system with $\X$ an $\mathcal{M}$-adhesive category and suppose that $\der{D}=\{\dder{D}_i\}_{i=0}^1$  is a derivation of length $2$ made by the following two weakly sequentially independent derivations $\dder{D}_0$, $\der{D}_1$, with independence pair $(i_0, i_1)$:
 	\[\xymatrix@C=15pt{L_0 \ar[d]_{m_0}&& K_0 \ar[d]_{k_0}\ar[ll]_{l_0} \ar[r]^{r_0} & R_0 \ar@/^.35cm/[drrr]|(.3)\hole_(.4){i_0} \ar[dr]|(.3)\hole_{h_0} && L_1 \ar@/_.35cm/[dlll]^(.4){i_1} \ar[dl]|(.3)\hole^{m_1}& K_1 \ar[d]^{k_1}\ar[l]_{l_1} \ar[rr]^{r_1} && R_1 \ar[d]^{h_1}\\G_0 && \ar[ll]^{f_0} D_0 \ar[rr]_{g_0}&& G_1  && \ar[ll]^{f_1} D_1 \ar[rr]_{g_1}&& G_2}\]

@@

\begin{definition} Let  $\dder{D}\colon H\Mapsto H$, $\dder{D}'\colon H\Mapsto T$ be two direct derivations in a left-linear DPO-rewriting system  $(\X, \R)$. An independence pair $(i_1, i_2)$ between $\dder{D}$ and $\dder{D}'$ is \emph{good} if there exists a filler $(u,u')$ between them such that  the squares below are pushouts.\[\xymatrix{K \ar[r]^{r} \ar[d]_{u}& R \ar[d]^{i_1} &K' \ar[r]^{l'} \ar[d]_{u'}& L'\ar[d]^{i_2} \\P \ar[r]_{p'}& D' & P \ar[r]_{p} & D}\]

@@

\begin{definition} Let $(\X, \R)$ be a left-linear DPO-rewriting system. Given two direct derivations $\dder{D}\colon G\Mapsto H$ and $\dder{D}'\colon H\Mapsto T$, we say that  $\dder{D}$ and $\dder{D}'$ are \emph{properly switchable} if $\dder{D}\updownarrow_! \dder{D'}$ and $S_{i_1,i_2}(\dder{D}')\updownarrow_!S_{i_1,i_2}(\dder{D})$, where $(i_2,i_2)$ is a good independence pair betwwen $\dder{D}$ and $\dder{D}'$. In such a case, we will write $\dder{D}\Updownarrow\dder{D}'$.

>>>>

\tableofcontents

\section*{Summary (with dependencies)}

\subsection*{REWRITING}
\begin{itemize}
\item DPO Rewriting in $\mathcal{M}$-adhesive categories, with left-$\mathcal{M}$-linear rules (from now on shortened in ``linear'')\\
  Observe that rewriting is deterministic (by uniqueness of
  PO-complement Corollary~\ref{lem:radj})

\item Category of abstract derivation for DPO-rewriting system $(\X, \R)$
  \begin{itemize}
  \item \emph{DPO-derivation category} $\dpo$: objects are graphs, arrows between possibly empty, derivations.
  \item \emph{DPO-abstract decorated derivation category} $\dpi$
  \end{itemize}

\item \emph{Consistent permutations} (Definition~\ref{def:permcon}):
  permutation on the steps of two abstract decorated derivations and isomorphism between the colimits such
  that arrows into the colimits ``commutes''.
  \begin{enumerate}
  \item It is unique for consuming rewriting systems
    (Corollary~\ref{cor:unique})
  \item They can be composed (Lemma~\ref{lem:sum})
  \end{enumerate}
\item

\item Notions of independence\\
  We have several possibile notions of independence for two direct
  derivations $\dder{D} ; \dder{D'}$
  \begin{enumerate}

  \item \emph{Existence of a filler}
    There is a filler for $\dder{D} ; \dder{D'}$
    (Definition~\ref{def:filler})

  \item \emph{Switchable}
    There is $\dder{D}_1' ; \dder{D}_1$ which is a \emph{switch}
    of $\dder{D} ; \dder{D'}$ (Definition~\ref{de:switch}, being a
    switch is defined axiomatically) [actually, Definition~\ref{de:good-switchable} provides a definition of switchable which looks different to me, I would avoid this, to be checked]

  \item \emph{(Weakly) Sequential independent} There is an
    \emph{independence pair}
    (Definition~\ref{de:sequential-independence}) They are called
    \emph{strongly sequential independent} if the independence pair is
    unique.
  \end{enumerate}

  We have that $(1) \Rightarrow (2)$ (Theorem~\ref{prop:fil}) and $(2) \Rightarrow (3)$ while
  we expect that in general $(3) \not\Rightarrow (1)$ (but no
  counterexample). This is instead known to hold for linear rules.

  Some terminology (Definition~\ref{de:good-switchable})
  \begin{itemize}

  \item \emph{Tame rewriting systems}\\
    An independence pairs induced by a filler is called \emph{good}
    A rewriting system where all indepedence pairs are good is called
    \emph{tame}

    NOTE: the categories in which condition $B$
    from~\cite{baldan2011adhesivity} identifies a large class of
    (quasi)adhesive categories where left-linear systems are all
    tame. Adapted here to M-adhesivity.

  \item \emph{Very Tame rewriting systems}\\
    Tame and existing indepence pairs are unique (hence switching is unique)
    (see later for examples)
  \end{itemize}

\item \emph{Switch equivalent derivations} (concrete)\\
  Definition~\ref{de:switchable}

  THEOREM: Two switch equivalent derivations admit a consistent
  permutation (Lemma~\ref{lem:consperm} ??).

  A consequence of this used for proving that the slice of traces is a pre-order.

  The converse holds only when rules are linear [proof clear, but not written]

\end{itemize}

\subsection*{CONCURRENCY}

This part is dealt with for very tame rewriting systems and depends on this assumption. Note that
\begin{itemize}
\item Linear rewriting systems on adhesive categories are very tame
\item Left-linear rewritint systems on graph-like categories with generalisation of (i) no isolated nodes in $L$ and (ii) right leg injective on nodes is very tame
\item Graphs with equivalences (Egg) with left-leg regular mono, right leg mono is very tame
\end{itemize}

We show
\begin{itemize}

\item Three-steps lemma (Lemma~\ref{lem:iig1}), which is a weakened version of ``indepedence is global''.
  \begin{itemize}

  \item If in $abc$ I have that $ab$ are independent and $c$ is anticipated, they remain independent.
  \item  If in $abc$ I have that $bc$ are independent and $a$ is moved to the end, they remain independent only if there is an additional condition (essentially requiring that $a$ was not a cause of $c$)
  \end{itemize}

\item No need of useless switchs (if two derivations are switch equivalent I can obtain one from another applying only inversions)

\item weak prime domain e connected es semantics
\end{itemize}

\section{Introduction}

@@

\begin{definition}[Sequential independence]
  \label{de:sequential-independence}
  Let $(\X, \R)$ be a left-linear DPO-rewriting system with $\X$ an $\mathcal{M}$-adhesive category. Let also $\dder{D}\colon G\Mapsto H$ and $\dder{D'}\colon H\Mapsto T$ be the two direct derivations depicted below.

@@

\begin{definition}[Switch]
  \label{de:switch}
 Let $(\X, \R)$ be a left-linear DPO-rewriting system with $\X$ an $\mathcal{M}$-adhesive category and suppose that $\der{D}=\{\dder{D}_i\}_{i=0}^1$  is a derivation of length $2$ made by the following two weakly sequentially independent derivations $\dder{D}_0$, $\der{D}_1$, with independence pair $(i_0, i_1)$:
 	\[\xymatrix@C=15pt{L_0 \ar[d]_{m_0}&& K_0 \ar[d]_{k_0}\ar[ll]_{l_0} \ar[r]^{r_0} & R_0 \ar@/^.35cm/[drrr]|(.3)\hole_(.4){i_0} \ar[dr]|(.3)\hole_{h_0} && L_1 \ar@/_.35cm/[dlll]^(.4){i_1} \ar[dl]|(.3)\hole^{m_1}& K_1 \ar[d]^{k_1}\ar[l]_{l_1} \ar[rr]^{r_1} && R_1 \ar[d]^{h_1}\\G_0 && \ar[ll]^{f_0} D_0 \ar[rr]_{g_0}&& G_1  && \ar[ll]^{f_1} D_1 \ar[rr]_{g_1}&& G_2}\]

@@

\begin{definition}
  \label{de:good-switchable}
  Let $\dder{D}\colon H\Mapsto H$, $\dder{D}'\colon H\Mapsto T$ be two
  direct derivations in a left-linear DPO-rewriting system $(\X,
  \R)$. An independence pair $(i_1, i_2)$ between $\dder{D}$ and
  $\dder{D}'$ is \emph{good} if there exists a filler $(u,u')$ between
  them such that the squares below are
  pushouts.\[\xymatrix{K \ar[r]^{r} \ar[d]_{u}& R \ar[d]^{i_1} &K'
      \ar[r]^{l'} \ar[d]_{u'}& L'\ar[d]^{i_2} \\P \ar[r]_{p'}& D' & P
      \ar[r]_{p} & D}\]

@@

\begin{definition}
  \label{de:switchable}
  Let $(\X, \R)$ be a left-linear DPO-rewriting system. Given two
  direct derivations $\dder{D}\colon G\Mapsto H$ and
  $\dder{D}'\colon H\Mapsto T$, we say that $\dder{D}$ and $\dder{D}'$
  are \emph{properly switchable} if $\dder{D}\updownarrow_! \dder{D'}$
  and $S_{i_1,i_2}(\dder{D}')\updownarrow_!S_{i_1,i_2}(\dder{D})$,
  where $(i_2,i_2)$ is a good independence pair betwwen $\dder{D}$ and
  $\dder{D}'$. In such a case, we will write
  $\dder{D}\Updownarrow\dder{D}'$.
